Rheology of Cubic Blue Phases

### 2012-11-02 papers/bprheo/bprheo.tex
(Partly) updated pictures
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1805 40876441-3496-0410-a613-a2fca35b2ae2

Our approach is based on the well-established Beris-Edwards model for hydrodynamics of
cholesteric liquid crystals \cite{Beris:1994}, which describes the ordered state 
in terms of a traceless, symmetric tensor order parameter ${\mathbf Q}({\mathbf r})$. 
In the uniaxial approximation, the order parameter is given by
$Q_{\alpha \beta}= q_s ( n_\alpha n_\beta - \frac{1}{3}\; \delta_{\alpha\beta})$
with ${\mathbf n}$ the director field and $q_s$ the amplitude of nematic
order. More generally,
the largest eigenvalue of ${\mathbf Q}$, $0\le q_s\le\frac{2}{3}$
characterises the local degree of orientational order.
The thermodynamic properties of the liquid crystal are determined by a free energy
${\cal F}$, whose density $f$ consists of a bulk contribution $f_b$ and a gradient part $f_g$, as follows,
\begin{eqnarray}
f_b&=&\frac{A_0}{2}\left(1-\frac{\gamma}{3}\right) Q_{\alpha \beta}^2\nonumber\\
&-&\frac{A_0 \gamma}{3}Q_{\alpha \beta} Q_{\beta \gamma} Q_{\gamma \alpha}+\frac{A_0 \gamma}{4}(Q_{\alpha \beta}^2)^2,\nonumber\\
f_g&=&\frac{K}{2}(\varepsilon_{\alpha\gamma\delta} \partial_\gamma Q_{\delta\beta}+2 q_0 Q_{\alpha \beta})^2+\frac{K}{2}(\partial_\beta Q_{\alpha \beta})^2.\label{FE}
\end{eqnarray}
The first term contains the bulk-free energy constant $A_0$ and the inverse temperature $\gamma$ which controls the magnitude of order.
The second part quantifies the cost of elastic distortions, which are proportional to the elastic constant $K$;
we work for simplicity in the one-elastic constant approximation. The wavevector $q_0=2\pi/p_0$, where $p_0$ is the cholesteric pitch.
The actual periodicity of the BP structure, $p$, does not need to be equal to $p_0$.
The ``redshift'' $r=p/p_0$ is adjusted during the simulation by following a simple procedure similar to~\cite{Alexander:2006}.

A particular thermodynamic state is specified by two dimensionless quantities: the effective temperature $\tau$ and chirality $\kappa$,
which are given by
\begin{eqnarray}
\tau&=&\frac{27(1-\gamma/3)}{\gamma}\nonumber\\
\kappa&=&\sqrt{\frac{108 K q_0^2}{A_0 \gamma}}\nonumber.
\end{eqnarray}
The dynamical evolution of the order parameter is given by the equation 
\begin{equation}
\left(\partial_t+ u_\alpha \partial_\alpha \right){\mathbf Q} - {\mathbf S}({\mathbf W},{\mathbf Q}) = \Gamma {\mathbf H}.
\label{op-eom}
\end{equation}
The first term on the left hand side of Eq.\ref{op-eom} is a material derivative, which describes the rate of change of a quantity moving along with the flow.
The second term accounts for the rate of change due to local velocity gradients $W_{\alpha \beta}=\partial_\beta u_\alpha$,
\begin{eqnarray}
{\mathbf S}({\mathbf W}, {\mathbf Q}) &=& (\xi {\mathbf A} + {\boldsymbol \Omega})({\mathbf Q}+\frac{\mathbf I}{3})\nonumber\\
& &\hspace*{-1.5cm}+ ({\mathbf Q}+\frac{\mathbf I}{3})(\xi {\mathbf A}  - {\boldsymbol \Omega})-2 \xi ({\mathbf Q}+\frac{\mathbf I}{3})
\mathrm{Tr}({\mathbf Q W}),
\label{sw}
\end{eqnarray}
where $\mathrm{Tr}$ denotes the tensorial trace, while 
${\mathbf A}=({\mathbf W}+{\mathbf W}^T)/2$ and
${\boldsymbol \Omega}=({\mathbf W}-{\mathbf W}^T)/2$ are the symmetric and antisymmetric part of the velocity gradient, respectively. $\xi$ 
is a constant depending on the molecular details of the liquid crystal.
Flow alignment occurs if $\xi \cos{2\theta}=(3q_s)/(2+q_s)$ has a real solution, where $\theta$ is the Leslie-angle: we select this case by 
setting $\xi=0.7$ in our simulations.
${\mathbf H}$ is the molecular field, which is a functional derivative of $\cal F$ that preserves the tracelessness of $\mathbf Q$:
\begin{equation}
{\bf H}=-\frac{\delta {\cal F}}{\delta {\bf Q}}+\frac{\bf I}{3}\,
\mathrm{Tr} \left(\frac{\delta {\cal F}}{\delta {\bf Q}}\right).
\label{molfield}
\end{equation}
The rotational diffusion constant $\Gamma$ in Eq.~\ref{op-eom} is proportional
to the inverse of the rotational viscosity $\gamma_1=2 q_s^2/\Gamma$
\cite{deGennes}.

The time evolution of the fluid density and velocity are respectively governed
by the continuity equation
$\partial_t \rho = -\partial_\alpha(\rho u_\alpha)$, and
the following Navier-Stokes equation:
\begin{eqnarray}
\partial_t u_\alpha +\rho \,u_\beta \partial_\beta u_\alpha
&=& \partial_\beta \Pi_{\alpha \beta}+ \eta\, \partial_\beta \{ \partial_\alpha u_\beta \nonumber\\
& &\hspace*{-1.5cm} +\; \partial_\beta u_\alpha
+(1+3\frac{\partial P_0}{\partial\rho} )\partial_\mu u_\mu \delta_{\alpha \beta}\}. 
\label{NSE}
\end{eqnarray}
The final term in Eq.~\ref{NSE} arises from the Chapman-Enskog expansion
of the LB equations \cite{Denniston:2001}.
At low flow rates the fluid can be considered as incompressible, so that the
last term on the right hand side of Eq.~\ref{NSE} remains small.
$\eta$ is an isotropic background viscosity which is set to $\eta=0.8333$ in LBU.
The pressure tensor reads explicitly
\begin{eqnarray}
\Pi_{\alpha \beta}&=&P_0 \delta_{\alpha\beta}
-\xi H_{\alpha \gamma}\left(Q_{\gamma \beta} +\frac{1}{3} \delta_{\gamma \beta}\right)\nonumber\\
&-&\xi \left(Q_{\alpha \gamma} +\frac{1}{3} \delta_{\alpha \gamma}\right) H_{\gamma \beta} + Q_{\alpha \gamma}H_{\gamma \beta}-H_{\alpha \gamma} Q_{\gamma \beta} \nonumber\\
&+&2 \xi  \left(Q_{\alpha \beta} +\frac{1}{3} \delta_{\alpha \beta}\right) Q_{\gamma \nu} H_{\gamma \nu}
- \partial_\alpha Q_{\gamma \nu} \frac{\delta{\cal F}}{\delta \partial_{\beta} Q_{\gamma \nu}}\nonumber\\
\label{Pi}
\end{eqnarray}
In the isotropic state ${\bf Q}\equiv 0$ and Eq.\ref{Pi} is reduced to the
scalar pressure which, in a system at rest, is constant to a very good
approximation.

The spectrum $X_\omega$ of the deviatoric stress $\Pi_{xy}(t)$ was determined via discrete Fourier transformation according to
\begin{equation}
X_\omega=\frac{1}{\sqrt{N}}\sum_{n=0}^{N-1} \left[\Pi_{xy}(t_n) - \bar{\Pi}_{xy}\right] \exp(2\pi i \, \omega \,n\,/\,N),
\label{spectrum}
\end{equation}
whereas we subtracted the average stress per cycle $\bar{\Pi}_{xy}$ from the signal.
$N$ was the number of points in the time series, typically one for every 100 LB time steps.

The system of coupled partial differential equations Eqs.~\ref{op-eom}
and~\ref{NSE} is solved by means of a
hybrid scheme~\cite{Marenduzzo:2007}. This uses a lattice Boltzmann algorithm
with predictor-corrector scheme for the continuity equation and
Eq.~\ref{NSE}, and a finite difference scheme for the equation of motion of
the tensor order parameter Eq.\ref{op-eom}. More details on the algorithm can
be found in~\cite{Denniston:2001, Denniston:2004}.

@@

In the following we report results of simulations of the bulk flow behaviour of the cubic blue 
phases BPI and BPII. 
The timestep and lattice spacing in lattice Boltzmann units (LBU) corresponds roughly to
$\sim 1 {\rm ns}$ and $\sim 10{\rm nm}$ in SI units. The LBU of stress
is equal to about $10^8$~Pa. Further details about the conversion 
from LBU to SI units can be found in ~\cite{Henrich:2011a,Henrich:2010b}.

The thermodynamic states are characterised by an effective temperature 
$\tau$ and the chirality $\kappa$, which measures the ratio of gradient to bulk free energy.
For the vast majority of simulations we opted for equilibrium states that are sufficiently 
far away from the cholesteric-isotropic transition, namely
$\tau=-0.5, \kappa=1.0$ in case of BPI and $\tau=-0.5, \kappa=2.0$ (BPII), respectively.
We performed also a number of simulations at metastable states at higher and lower temperatures 
and different at different chiralities, but didn't find any significant differences in the 
generic flow behaviour that is described below.
However, for thermodynamic states that are closer to the phase boundary the disclination 
network broke up into non-BP states at lower shear rates. This is not surprising as
these states have on average higher free energy densities and smaller order parameters
than those specified above and can therefore not withstand the same external forces.

@@

Simple shear flow was imposed by means of a Lees-Edwards boundary condition ~\cite{Wagner:2002} with
the top (bottom) half of the planes moving in positive (negative) x-direction and the 
velocity gradient along the y-direction.
We applied a correction for spurious momentum currents that emerge due to required interpolations across 
the Lees-Edwards planes \cite{Henrich:2012a}.
The shear rates were varied over more than two orders of magnitude from about 
$\gd=2.44\times \e{-6}$ to $6.25\times\e{-4}$ in LB units.
For clarity we classify various sub-regimes, two in case of BPII and three in case of BPI. 
They are BPII-1 ($\gd < 6.25\e{-4}$), BPII-2 ($\gd\ge 6.25\e{-4}$) and
BPI-1 ($\gd < 3.91\e{-5}$), BPI-2 ($3.91\e{-5}\le\gd< 6.25\e{-4}$) and BPI-3 ($\gd\ge 6.25\e{-4}$). 
Disclination lines are plotted as isosurface of the scalar order parameter $q$. Typical values 
are $q_s=0.18$ for BPI and $q_s=0.15$ for BPII.

@@

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{stress_bp2.pdf}
\includegraphics[width=0.495\textwidth]{stress_vs_strain_bp2.pdf}
\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\eta\,\gd$ of BPII over time (top) and strain (bottom). 
For clarity the individual curves have been shifted along the time and strain axis. 
The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$ and two different regimes of flow: 
regular periodic breakup and reconnecting of the disclination lines (BPII-1, open circles) 
and break up of the network into a travelling helical wave (BPII-2, solid circle).}
\label{bp2-rheo}
\end{figure}

We start our discussion with BPII as its flow behaviour turns out to be somewhat simpler 
than that of BPI. BPII has simple cubic symmetry and the $\pi$-disclination lines intersect and form a characteristic network of nodes.
Ahead of the more detailed discussion we provide first a general overview of the flow behaviour at all applied shear rates.
Fig. \ref{bp2-rheo} shows the apparent viscosity, which is defined as the 
deviatoric stress normalised to the isotropic background viscosity: 
$\eta_{app}=\Pi_{xy}/\eta\,\gd$.
A numerical value of $\eta_{app}=1$ corresponds to total shear thinning without 
any additional contribution from the liquid crystal.
The top picture shows the date over time, whereas the graph at the bottom gives 
the same data versus accumulated strain.
For all but the highest flow rates $\eta_{app}$ oscillates sinusoidally, which
is entirely because of the periodic breakup and reconnecting of the network in shear flow. We refer to this regime as BPII-1.

The inset shows a kind of flow curve, defined as time averages of 
the deviatoric stress tensor $\bar{Pi}_{xy}$ as a function of shear rate $\gd$.
The error bars indicate the maximum and minimum stresses that occur during one cycle.
For all but the largest shear rates a power law fit $\bar{\Pi}_{xy}=a \gd^b$ with 
$a=0.35, b=0.95$ describes the data to a very good approximation. 
Hence, the degree of shear-thinning is remarkably small, but may depend
on the exact direction in which the strain is applied, an aspect that we     
leave for future work.

@@

\begin{figure}[htpb]
\includegraphics[width=0.4\textwidth]{disc-160k_run902.png}
\includegraphics[width=0.4\textwidth]{disc-164k_run902.png}
\includegraphics[width=0.4\textwidth]{disc-165k_run902.png}
\caption{Disclination network of BPII in shear flow: 
The pictures show a typical sequence of snapshots in the steady state 
at $\gd=1.56\e{-4}$ and time steps $t=1.60, 1.64,1.65\e{5}$. The velocity 
gradient is oriented along the vertical direction, whereas the 
horizontal direction is the flow direction. Lees-Edwards boundary 
conditions have been imposed in such a way that network moves to the 
right in the upper half and to the left in the lower half of the picture.}
\label{bp2-med}
\end{figure}

\begin{figure*}[htpb]
\includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run902.png}
\includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run903.png}
\caption{Velocity patterns and disclination network in BPII for positive (left) and negative (right) helicity of the 
underlying cholesteric helix: The pictures show velocity vectors $(0,v_y,v_z)$ in x-direction, 
i.e. the velocity has been projected onto a plane perpendicular to the flow direction. 
The vertical and horizontal direction are the gradient and vorticity direction, respectively.
The colour code gives the magnitude and sign of the component in vorticity direction.
The snapshot shows a typical frame during a periodically recurring sequence.
The network on the left with positive helicity travels rightwards, whereas the one one the right
moves leftwards. The recurrence period in vorticity direction is six times longer than the time it takes the network to reconnect in flow direction.}
\label{bp2-velo}
\end{figure*}

@@

Interestingly, while being displaced with the flow the entire network moves perpendicular 
to the flow and velocity gradient in the vorticity direction.
A similar behaviour has been recently observed for blue phases in shear flow and 
confined geometries ~\cite{Henrich:2012b}.
But contrary to the stick-slip motion that has been reported there the movement is 
steady. It occurs in such a manner that the positions of breakup and reconnection 
of the network, visible in Fig. \ref{bp2-med}, are slightly offset and make the network 
travel along the z-direction.
The recurrence time in vorticity direction $\tau_V=6\tau_F$, 
i.e. it takes a displacement of six unit cells in flow directions 
until the network has moved one unit cell in vorticity direction.

Fig. \ref{bp2-velo} shows the disclination network sometime between reconnecting 
and the next breakup with superimposed velocity vectors.
Shown are what we may call secondary velocity components, obtained by projecting the
velocity onto a plain perpendicular to the flow direction.
This allows to get rid of the dominating velocity component $v_x$ and to
visualise the two much smaller components $v_y$ and $v_z$.
The magnitude of the secondary components is typically in the range of a 
few percent of the primary flow component.

Characteristic bands are visible, which are oriented along the vorticity direction and
normally hidden behind the primary flow component.
When the helicity of the underlying cholesteric phase is changed from left- to a 
right-handed the pattern becomes a mirror image of the former one. The sign of 
$v_z$ change and also the sense of motion is inverted.
Further quantitative evidence for a direct link between the sense of motion 
and the helicity can be gained by time-averaging over individual cycles.
Tab. \ref{tab1} gives minima, maxima, averages and standard deviations 
of the velocity components.
All values for the two runs with inverted helicity are identical apart 
from a change of sign in the z-components.
There is only a slight imbalance between the maximum and minimum velocities 
in z-direction, suggesting that the movement of the network in vorticity 
direction is permeative. Note that this imbalance is retained under inversion of
helicity.

@@

In our scenario the break-up occurs for shear rates $\gd>3.12\e{-4}$, which
corresponds to Ericksen numbers of ${\cal E}\simeq10$, based on our
parameters ($K=0.02, \eta=0.66$), typical velocities (viz. Tab. \ref{tab1}) 
and length scales ($p/2$).

@@

BPI has body-centered symmetry and contrary to BPII the disclination lines
 are well separated and do not intersect.
We believe that this topological difference is responsible for most of
the differences between BPI and BPII regarding their rheological response.

We present again key aspects in a kind of overview before we 
address some features in more detail. Fig. \ref{bp1-rheo} shows the 
apparent viscosity $\eta_{app}$ over time and strain for the same shear rates
as in Fig. \ref{bp2-rheo}.
For shear rates in regime BPI-3, i.e. $\gd\ge6.25\e{-4}$ a breakup of 
the disclination network occurs similar to and at the same critical shear rate as 
the one observed in BPII-2 but with an order structure that is a quasi 
two-dimensional.

In regime BPI-2 at lower shear rates periodic oscillations emerge in the 
deviatoric stress that are not sinusoidal, but show an increasingly complex
pattern up to the breakup point.
For even lower shear rates $\gd\le3.91\e{-5}$ in regime BPI-1 
they become  irregularunregular and finally disappear after a few
cycles. Shear-thinning is slightly stronger than in BPII. 
The average shear stress during one cycle can be fitted reasonably 
well by a power law $\bar{\Pi}_{xy}=a \gd^b$ with $a=0.019, b=0.63$.

>>>>

Our approach is based on the well-established Beris-Edwards model for hydrodynamics of
cholesteric liquid crystals \cite{Beris:1994}, which describes the ordered state 
in terms of a traceless, symmetric tensor order parameter ${\mathbf Q}({\mathbf r})$. 
In the uniaxial approximation, the order parameter is given by
$Q_{\alpha \beta}= q_s ( \hat{n}_\alpha \hat{n}_\beta - \frac{1}{3}\; \delta_{\alpha\beta})$
with $\hat{{\mathbf n}}$ the director field and $q_s$ the amplitude of nematic
order. More generally,
the largest eigenvalue of ${\mathbf Q}$, $0\le q_s\le\frac{2}{3}$
characterises the local degree of orientational order.
The thermodynamic properties of the liquid crystal are determined by a free energy
${\cal F}$, whose density $f$ consists of a bulk contribution $f_b$ and a gradient part $f_g$, as follows,
\begin{eqnarray}
f_b&=&\frac{A_0}{2}\left(1-\frac{\gamma}{3}\right) Q_{\alpha \beta}^2\nonumber\\
&-&\frac{A_0 \gamma}{3}Q_{\alpha \beta} Q_{\beta \gamma} Q_{\gamma \alpha}+\frac{A_0 \gamma}{4}(Q_{\alpha \beta}^2)^2,\nonumber\\
f_g&=&\frac{K}{2}(\varepsilon_{\alpha\gamma\delta} \partial_\gamma Q_{\delta\beta}+2 q_0 Q_{\alpha \beta})^2+\frac{K}{2}(\partial_\beta Q_{\alpha \beta})^2.\label{FE}
\end{eqnarray}
The first term contains the bulk-free energy constant $A_0$ and the inverse temperature $\gamma$ which controls the magnitude of order.
The second part quantifies the cost of elastic distortions, which are proportional to the elastic constant $K$;
we work for simplicity in the one-elastic constant approximation. The wavevector $q_0=2\pi/p_0$, where $p_0$ is the cholesteric pitch.
The actual periodicity of the BP structure, $p$, does not need to be equal to $p_0$.
The ``redshift'' $r=p/p_0$ is adjusted during the simulation by following a simple procedure~\cite{Alexander:2006}.

A particular thermodynamic state is specified by two dimensionless quantities: the effective temperature 
\begin{equation}
\tau=\frac{27(1-\gamma/3)}{\gamma}
\end{equation}
and chirality 
\begin{equation}
\kappa=\sqrt{\frac{108 K q_0^2}{A_0 \gamma}},
\end{equation}
which measures the ratio of gradient to bulk free energy.

The dynamical evolution of the order parameter is given by the equation 
\begin{equation}
\left(\partial_t+ v_\alpha \partial_\alpha \right){\mathbf Q} - {\mathbf S}({\mathbf W},{\mathbf Q}) = \Gamma {\mathbf H}.
\label{op-eom}
\end{equation}
The first term on the left hand side of Eq.\ref{op-eom} is a material derivative, which describes the rate of change of a quantity moving along with the flow.
The second term accounts for the rate of change due to local velocity gradients $W_{\alpha \beta}=\partial_\beta v_\alpha$,
\begin{eqnarray}
{\mathbf S}({\mathbf W}, {\mathbf Q}) &=& (\xi {\mathbf A} + {\boldsymbol \Omega})({\mathbf Q}+\frac{\mathbf I}{3})\nonumber\\
& &\hspace*{-1.5cm}+ ({\mathbf Q}+\frac{\mathbf I}{3})(\xi {\mathbf A}  - {\boldsymbol \Omega})-2 \xi ({\mathbf Q}+\frac{\mathbf I}{3})
\mathrm{Tr}({\mathbf Q W}),
\label{sw}
\end{eqnarray}
where $\mathrm{Tr}$ denotes the tensorial trace, while 
${\mathbf A}=({\mathbf W}+{\mathbf W}^T)/2$ and
${\boldsymbol \Omega}=({\mathbf W}-{\mathbf W}^T)/2$ are the symmetric and antisymmetric part of the velocity gradient, respectively. $\xi$ 
is a constant depending on the molecular details of the liquid crystal.
Flow alignment occurs if $\xi \cos{2\theta}=(3q_s)/(2+q_s)$ has a real solution, where $\theta$ is the Leslie-angle: we select this case by 
setting $\xi=0.7$ in our simulations.
${\mathbf H}$ is the molecular field, which is the functional derivative of $\cal F$ that preserves the tracelessness of $\mathbf Q$:
\begin{equation}
{\bf H}=-\frac{\delta {\cal F}}{\delta {\bf Q}}+\frac{\bf I}{3}\,
\mathrm{Tr} \left(\frac{\delta {\cal F}}{\delta {\bf Q}}\right).
\label{molfield}
\end{equation}
The rotational diffusion constant $\Gamma$ in Eq.~\ref{op-eom} is proportional
to the inverse of the rotational viscosity $\gamma_1=2 q_s^2/\Gamma$
\cite{deGennes}.

The time evolution of the fluid density and velocity are respectively governed
by the continuity equation
$\partial_t \rho = -\partial_\alpha(\rho v_\alpha)$, and
the following Navier-Stokes equation:
\begin{eqnarray}
\partial_t v_\alpha +\rho \,v_\beta \partial_\beta v_\alpha
&=& \partial_\beta \Pi_{\alpha \beta}+ \eta\, \partial_\beta \{ \partial_\alpha v_\beta \nonumber\\
& &\hspace*{-1.5cm} +\; \partial_\beta v_\alpha
+(1+3\frac{\partial P_0}{\partial\rho} )\partial_\mu v_\mu \delta_{\alpha \beta}\}. 
\label{NSE}
\end{eqnarray}
The final term in Eq.~\ref{NSE} arises from the Chapman-Enskog expansion
of the Lattice Boltzmann (LB) equations \cite{Denniston:2001}.
At low flow rates the fluid can be considered as incompressible, so that the
last term on the right hand side of Eq.~\ref{NSE} remains small.
$\eta$ is an isotropic background viscosity which is set to $\eta=0.8333$ in LB units.
The pressure tensor reads explicitly
\begin{eqnarray}
\Pi_{\alpha \beta}&=&P_0 \delta_{\alpha\beta}
-\xi H_{\alpha \gamma}\left(Q_{\gamma \beta} +\frac{1}{3} \delta_{\gamma \beta}\right)\nonumber\\
&-&\xi \left(Q_{\alpha \gamma} +\frac{1}{3} \delta_{\alpha \gamma}\right) H_{\gamma \beta} + Q_{\alpha \gamma}H_{\gamma \beta}-H_{\alpha \gamma} Q_{\gamma \beta} \nonumber\\
&+&2 \xi  \left(Q_{\alpha \beta} +\frac{1}{3} \delta_{\alpha \beta}\right) Q_{\gamma \nu} H_{\gamma \nu}
- \partial_\alpha Q_{\gamma \nu} \frac{\delta{\cal F}}{\delta \partial_{\beta} Q_{\gamma \nu}}\nonumber\\
\label{Pi}
\end{eqnarray}
In the isotropic state ${\bf Q}\equiv 0$ and Eq.\ref{Pi} is reduced to the
scalar pressure which, in a system at rest, is constant to a very good
approximation.

It is possible to define a dimensionless number that describes the deformation
of the director field under flow. The so-called Ericksen number
is defined as 
\begin{equation}
Er=\frac{\eta V L}{K}
\end{equation}
with $\eta$ and $K$ being the viscosity and elastic constant and
$V$ and $L$ as typical velocity and length scale.

The spectrum $X_\omega$ of the deviatoric stress $\Pi_{xy}(t)$ was determined via discrete Fourier transformation according to
\begin{equation}
X_\omega=\frac{1}{\sqrt{N}}\sum_{n=0}^{N-1} \left[\Pi_{xy}(t_n) - \bar{\Pi}_{xy}\right] \exp(2\pi i \, \omega \,n\,/\,N),
\label{spectrum}
\end{equation}
where we subtract the average stress per cycle $\bar{\Pi}_{xy}$ from the signal.
$N$ was the number of points in the time series, typically one for every 100 LB time steps.

The system of coupled partial differential equations Eqs.~\ref{op-eom}
and~\ref{NSE} is solved by means of a
hybrid scheme~\cite{Marenduzzo:2007}. 
More details on the algorithm can
be found in~\cite{Denniston:2001, Denniston:2004}.

@@

In the following we report results of simulations of the bulk flow behaviour of the cubic blue 
phases BPI and BPII. 
The timestep and lattice spacing in lattice Boltzmann units (LBU) corresponds roughly to
$\sim 1 {\rm ns}$ and $\sim 10{\rm nm}$ in SI units. The LB unit of stress
is equal to about $10^8$~Pa. Further details about the conversion 
from LBU to SI units can be found in ~\cite{Henrich:2011a,Henrich:2010b}.

For the vast majority of simulations we opted for thermodynamic states that are 
well inside the equilibrium region of the individual blue phase and far away 
from the cholesteric-isotropic transition. Temperature and chirality were 
$\tau=-0.5, \kappa=1.0$ in case of BPI and $\tau=-0.5, \kappa=2.0$ for BPII, respectively.
For these parameters the total free energy density $f$ remained always negative, 
even at the highest flow rate and despite larger gradient contributions $f_g$.
We performed also a number of simulations at metastable states at higher and lower temperatures 
and different at different chiralities, but didn't find any significant differences in the 
general flow behaviour that is described below.
However, for thermodynamic states that are closer to the phase boundary the disclination 
network broke up into non-BP states at lower shear rates. This is not surprising as
these states have on average higher free energy densities and smaller order parameters
than those specified above and can therefore not withstand the same external forces.

@@

Simple shear flow was imposed by means of a Lees-Edwards boundary condition ~\cite{Wagner:2002} with
the top (bottom) part of the system flowing in positive (negative) x-direction and the 
velocity gradient along the y-direction.
The shear rates were varied over more than two orders of magnitude from about 
$\gd=2.44\times \e{-6}$ to $6.25\times\e{-4}$ in LB units.
For clarity we classify various sub-regimes, two in case of BPII and three in case of BPI. 
They are BPII-1 ($\gd < 6.25\e{-4}$), BPII-2 ($\gd\ge 6.25\e{-4}$) and
BPI-1 ($\gd < 3.91\e{-5}$), BPI-2 ($3.91\e{-5}\le\gd< 6.25\e{-4}$) and BPI-3 ($\gd\ge 6.25\e{-4}$). 
Disclination lines are plotted as isosurface of the scalar order parameter $q$. Typical values 
are $q_s=0.18$ for BPI and $q_s=0.15$ for BPII.

@@

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{stress_bp2.pdf}
\includegraphics[width=0.495\textwidth]{stress_vs_strain_bp2.pdf}
\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\eta\,\gd$ of BPII versus time (top) and strain (bottom). 
For clarity the individual curves have been shifted along the time and strain axis. 
The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$ and two different regimes of flow: 
regular periodic breakup and reconnecting of the disclination lines (BPII-1, open circles) 
and break up of the network into a travelling helical wave (BPII-2, solid circle).}
The error bars indicate the maximum and minimum stresses that occur during one cycle.
\label{bp2-rheo}
\end{figure}

We start our discussion with BPII as its flow behaviour turns out to be somewhat simpler 
than that of BPI. BPII has simple cubic symmetry and the disclination lines intersect and form a characteristic network of nodes.
Ahead of the more detailed discussion we provide first a general overview of the flow behaviour at all applied shear rates.
Fig. \ref{bp2-rheo} shows the apparent viscosity, which is defined as the 
deviatoric stress normalised to the isotropic background viscosity: 
$\eta_{app}=\Pi_{xy}/\eta\,\gd$.
A numerical value of $\eta_{app}=1$ corresponds to total shear thinning without 
any additional contribution from the liquid crystal.
The top picture shows the data over time, whereas the graph at the bottom gives 
the same data versus accumulated strain.
For all but the highest flow rates $\eta_{app}$ oscillates sinusoidally, which
is because of the periodic breakup and reconnecting of the network in shear flow. We refer to this regime as BPII-1.

The inset shows a flow curve, defined as time averages of 
the deviatoric stress tensor $\bar{\Pi}_{xy}$ as a function of shear rate $\gd$.
For all but the largest shear rates a power law fit $\bar{\Pi}_{xy}=a \gd^b$ with 
$a=0.35, b=0.95$ describes the data to a very good approximation. 
Hence, the degree of shear-thinning is remarkably small, but may depend
on the exact direction in which the strain is applied, an aspect that we     
leave for future work.

@@

\begin{figure}[htpb]
\includegraphics[width=0.41\textwidth]{disc-160k_run902.png}
\includegraphics[width=0.41\textwidth]{disc-164k_run902.png}
\includegraphics[width=0.41\textwidth]{disc-165k_run902.png}
\caption{Disclination network of BPII in shear flow: 
The pictures show a typical sequence of snapshots in the steady state 
at $\gd=1.56\e{-4}$ and time steps $t=1.60, 1.64,1.65\e{5}$. The velocity 
gradient is oriented along the vertical direction, whereas the 
horizontal direction is the flow direction. Lees-Edwards boundary 
conditions have been imposed in such a way that network moves to the 
right in the upper half and to the left in the lower half of the picture.}
\label{bp2-med}
\end{figure}

\begin{figure*}[htpb]
\includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run902.png}
\includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run903.png}
\caption{Velocity patterns and disclination network in BPII for positive (left) and negative (right) helicity of the 
underlying cholesteric helix: The pictures show velocity vectors $(0,v_y,v_z)$.
The view is in x-direction. This means the velocity ${\mathbf v}$ 
has been projected onto a plane perpendicular to the flow direction. 
The vertical and horizontal direction are the gradient and vorticity direction, respectively.
The colour code gives the magnitude and sign of the component in vorticity direction.
The snapshot shows a typical frame during a periodically recurring sequence.
The network on the left with positive helicity travels rightwards, whereas the one one the right
moves leftwards. The recurrence period in vorticity direction is six times longer than the time it takes the network to reconnect in flow direction.}
\label{bp2-velo}
\end{figure*}

@@

Interestingly, while being displaced with the flow the entire network moves 
in the vorticity direction.  A similar behaviour has been recently 
observed for blue phases in shear flow in 
confined geometries ~\cite{Henrich:2012b}.
In contrast to the stick-slip motion that has been reported there the movement is 
steady. It occurs in such a manner that the positions of breakup and reconnection 
of the network, visible in Fig. \ref{bp2-med}, are slightly offset and make the network 
travel along the z-direction.
The recurrence time in vorticity direction $\tau_V=6\tau_F$, 
i.e. it takes a displacement of six unit cells in flow directions 
until the network has moved one unit cell in vorticity direction.

Fig. \ref{bp2-velo} shows the disclination network between reconnecting 
and the next breakup with superimposed velocity vectors.
Shown are what we may call secondary velocity components, obtained by projecting the
velocity onto a plane perpendicular to the flow direction.
This gets rid of the dominating velocity component $v_x$ and 
allows to visualise the two much smaller components $v_y$ and $v_z$.
The magnitude of the secondary components is typically in the range of a 
few percent of the primary flow component.

Characteristic bands are visible in Fig. \ref{bp2-velo}, which are 
oriented along the vorticity direction. They are obscured by the
primary flow component if the above described projection procedure
is not applied.
When the helicity of the underlying cholesteric phase is changed from 
left-handed to right-handed, 
the pattern becomes a mirror image of the former one. The sign of 
$v_z$ changes and the sense of motion is inverted.
Further quantitative evidence for a direct link between the sense of motion 
and the helicity can be gained by time-averaging over individual cycles.
Tab. \ref{tab1} gives minima, maxima, averages and standard deviations 
of the velocity components.
All values for the two runs with inverted helicity are identical apart 
from a change of sign in the z-components.
There is only a slight imbalance between the maximum and minimum velocities 
in z-direction, suggesting that the movement of the network in vorticity 
direction is permeative. Note that this imbalance is retained under inversion of
helicity.

@@

In our scenario the break-up occurs for shear rates $\gd>3.12\e{-4}$, which
corresponds to Ericksen numbers of ${\it Er}\simeq10$, based on our
parameters ($K=0.02, \eta=0.66$), typical velocities (viz. Tab. \ref{tab1}) 
and length scales ($p/2$).

@@

BPI has body-centered symmetry and unlike BPII. The disclination lines
 are well separated and do not intersect.
We believe that this topological difference is responsible for most of
the differences between BPI and BPII regarding their rheological response.

We present again key aspects in an overview before we 
address some features in more detail. Fig. \ref{bp1-rheo} shows the 
apparent viscosity $\eta_{app}$ versus time and strain for the same shear rates
as in Fig. \ref{bp2-rheo}.
For shear rates in regime BPI-3, i.e. $\gd\ge6.25\e{-4}$ a breakup of 
the disclination network occurs. This is the same critical shear rate as 
the one observed in BPII-2. But this time the configuration is quasi 
two-dimensional.

In regime BPI-2 at lower shear rates periodic oscillations emerge in the 
deviatoric stress that are not sinusoidal, but show an increasingly complex
pattern up to the breakup point.
For even lower shear rates $\gd\le3.91\e{-5}$ in regime BPI-1 
these oscillations become very erratic. There is no regular recurrence and 
after a few cycles the stress pattern is completely irregular. 
Shear-thinning is slightly stronger than in BPII. 
The average shear stress during one cycle can be fitted reasonably 
well by a power law $\bar{\Pi}_{xy}=a \gd^b$ with $a=0.019, b=0.63$.

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{stress_bp1.pdf}
\includegraphics[width=0.495\textwidth]{stress_vs_strain_bp1.pdf}
\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\dot{\gamma}$ of BPI versus time (top) 
and strain (bottom). The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$. 
Depending on the configuration in steady shear flow three different regimes 
can be identified: amorphous BP network with yield stress (BPI-1, triangles), 
steady flow with recurring patterns (BPI-2, open circles) and 
breakup of the disclination network (BPI-3, solid circle). 
The error bars represent the minimum and maximum shear stress 
during one cycle. For clarity the individual curves have been shifted along
the time and strain axis.}
\label{bp1-rheo}
\end{figure}
